Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-16 jss/cheem_paper.Rmd
figure captions fixed

(ref:casepenguins-cap) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a) and the dashed vertical line in (b). The radial view shows varying the contribution of `fl` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \url{https://vimeo.com/666431172}.)

```{r casepenguins, echo=F, out.width="100%", fig.cap = '(ref:casepenguins-cap)'}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

(ref:casepenguinsblfl-cap) Checking what is learned from the cheem viewer. This is a plot of flipper length (`fl`) and bill length (`bl`), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small `fl` length which is why it is confused with a Chinstrap.

```{r casepenguinsblfl, echo=F, out.width="100%", fig.cap = '(ref:casepenguinsblfl-cap)'}
knitr::include_graphics("./figures/case_penguins_BlFl.png")
```

@@

(ref:casechocolates-cap) Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this observation. So, the contribution of Sugar is varied---reducing it corresponds primarily with an increased magnitude from Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at \url{https://vimeo.com/666431143}.)

```{r casechocolates, echo=F, out.width="100%", fig.cap = '(ref:casechocolates-cap)'}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

(ref:casechocolatesinverse-cap) Examining the local explanation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed variable importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at \url{https://vimeo.com/666431148}.)

```{r casechocolatesinverse, echo=F, out.width="100%", fig.cap = '(ref:casechocolatesinverse-cap)'}
knitr::include_graphics("./figures/case_chocolates_inverse.png")
```

@@

(ref:casefifa-cap) Exploring the wages relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI, and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown on the left, and it can be seen that this combination of variables produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen variable to vary. It starts very low, and Messi's predicted wages decrease dramatically as its contribution increases (right plot). The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at \url{https://vimeo.com/666431163}.)

```{r casefifa, echo=F, out.width = "90%", fig.cap = '(ref:casefifa-cap)'}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

(ref:caseames-cap) Exploring an observation with an extreme residual as the PI in relation to an observation with an accurate prediction for a similarly priced house in a random forest fit to the Ames housing data. The local explanation indicates a sizable attribution to Lot Area (`LtA`), while the CI has minimal attribution to this variable. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \url{https://vimeo.com/666431134}.)
```{r caseames, echo=F, out.width="90%", fig.cap = '(ref:caseames-cap)'}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

An implementation is provided in the open-source __R__ package __cheem__, available on CRAN at \url{https://CRAN.R-project.org/package=cheem}. Example data sets are provided, and you can upload your data after model fitting and computing the local explanations. The local explanations need to be pre-computed and uploaded. Examples show how to do this for tree SHAP values, using  __treeshap__ (tree based models from __gbm__, __lightgbm__, __randomForest__, __ranger__, or __xgboost__  @greenwell_gbm_2020; @shi_lightgbm_2022; @liaw_classification_2002; @wright_ranger_2017; @chen_xgboost_2021, respectively). The SHAP and oscillation explanations could be easily added using `DALEX::explain()` [@biecek_dalex_2018; @biecek_explanatory_2021].

@@

- A version of the cheem viewer shiny app can be directly accessed at
\url{https://ebsmonash.shinyapps.io/cheem_initial/}.
- The development version of the package is available at \url{https://github.com/nspyrison/cheem}, and
- Documentation of the package can be found at \url{https://nspyrison.github.io/cheem/}.

>>>>

```{r casepenguins, echo=F, out.width="100%", fig.cap = "Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a) and the dashed vertical line in (b). The radial view shows varying the contribution of `fl` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at https://vimeo.com/666431172.)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

```{r casepenguinsblfl, echo=F, out.width="100%", fig.align='center', fig.cap = "Checking what is learned from the cheem viewer. This is a plot of flipper length (`fl`) and bill length (`bl`), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small `fl` length which is why it is confused with a Chinstrap."}
knitr::include_graphics("./figures/case_penguins_BlFl.png")
```

@@

```{r casechocolates, echo=F, out.width="100%", fig.cap = "Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this observation. So, the contribution of Sugar is varied---reducing it corresponds primarily with an increased magnitude from Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at https://vimeo.com/666431143.)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

```{r casechocolatesinverse, echo=F, out.width="100%", fig.cap = "Examining the local explanation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed variable importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at https://vimeo.com/666431148.)"}
knitr::include_graphics("./figures/case_chocolates_inverse.png")
```

@@

```{r casefifa, echo=F, out.width = "90%", fig.cap = "Exploring the wages relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI, and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown on the left, and it can be seen that this combination of variables produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen variable to vary. It starts very low, and Messi's predicted wages decrease dramatically as its contribution increases (right plot). The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at https://vimeo.com/666431163.)"}
knitr::include_graphics("./figures/case_fifa.png")
```

@@

```{r caseames, echo=F, out.width="90%", fig.cap = "Exploring an observation with an extreme residual as the PI in relation to an observation with an accurate prediction for a similarly priced house in a random forest fit to the Ames housing data. The local explanation indicates a sizable attribution to Lot Area (`LtA`), while the CI has minimal attribution to this variable. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at https://vimeo.com/666431134.)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

@@

An implementation is provided in the open-source __R__ package __cheem__, available on CRAN at https://CRAN.R-project.org/package=cheem. Example data sets are provided, and you can upload your data after model fitting and computing the local explanations. The local explanations need to be pre-computed and uploaded. Examples show how to do this for tree SHAP values, using  __treeshap__ (tree based models from __gbm__, __lightgbm__, __randomForest__, __ranger__, or __xgboost__  @greenwell_gbm_2020; @shi_lightgbm_2022; @liaw_classification_2002; @wright_ranger_2017; @chen_xgboost_2021, respectively). The SHAP and oscillation explanations could be easily added using `DALEX::explain()` [@biecek_dalex_2018; @biecek_explanatory_2021].

@@

- A version of the cheem viewer shiny app can be directly accessed at
https://ebsmonash.shinyapps.io/cheem_initial/.
- The development version of the package is available at https://github.com/nspyrison/cheem, and
- Documentation of the package can be found at https://nspyrison.github.io/cheem/.

### 2023-05-16 jss/cheem_paper.tex
figure captions fixed

(ref:casepenguins-cap) Examining the SHAP values for a random forest
model classifying Palmer penguin species. The PI is a Gentoo (purple)
penguin that is misclassified as a Chinstrap (orange), marked as an
asterisk in (a) and the dashed vertical line in (b). The radial view
shows varying the contribution of \texttt{fl} from the initial
attribution projection (b, left), which produces a linear combination
where the PI is more probably (higher density value) a Chinstrap than a
Gentoo (b, right). (The animation of the radial tour is at
\url{https://vimeo.com/666431172}.)

@@

\caption[(ref:casepenguins-cap)]{(ref:casepenguins-cap)}\label{fig:casepenguins}
\end{figure}
\end{CodeChunk}

@@

(ref:casepenguinsblfl-cap) Checking what is learned from the cheem
viewer. This is a plot of flipper length (\texttt{fl}) and bill length
(\texttt{bl}), where an asterisk highlights the PI. A Gentoo (purple)
misclassified as a Chinstrap (orange). The PI has an unusually small
\texttt{fl} length which is why it is confused with a Chinstrap.

@@

\caption[(ref:casepenguinsblfl-cap)]{(ref:casepenguinsblfl-cap)}\label{fig:casepenguinsblfl}
\end{figure}
\end{CodeChunk}

@@

(ref:casechocolates-cap) Examining the local explanation for a PI which
is dark (orange) chocolate incorrectly predicted to be milk (green).
From the attribution projection, this chocolate correctly looks more
like dark than milk, which suggests that the local explanation does not
help understand the prediction for this observation. So, the
contribution of Sugar is varied---reducing it corresponds primarily with
an increased magnitude from Fiber. When Sugar is zero, Fiber contributes
strongly towards the left. In this view, the PI is closer to the bulk of
the milk chocolates, suggesting that the prediction put a lot of
importance on Fiber. This chocolate is a rare dark chocolate without any
Fiber leading to it being mistaken for a milk chocolate. (A video of the
tour animation can be found at \url{https://vimeo.com/666431143}.)

@@

\caption[(ref:casechocolates-cap)]{(ref:casechocolates-cap)}\label{fig:casechocolates}
\end{figure}
\end{CodeChunk}

@@

(ref:casechocolatesinverse-cap) Examining the local explanation for a PI
which is milk (green) chocolate incorrectly predicted to be dark
(orange). In the attribution projection, the PI could be either milk or
dark. Sodium and Fiber have the largest differences in attributed
variable importance, with low values relative to other milk chocolates.
The lack of importance attributed to these variables is suspected of
contributing to the mistake, so the contribution of Sodium is varied. If
Sodium had a larger contribution to the prediction (like in this view).
the PI would look more like other milk chocolates. (A video of the tour
animation can be found at \url{https://vimeo.com/666431148}.)

@@

\caption[(ref:casechocolatesinverse-cap)]{(ref:casechocolatesinverse-cap)}\label{fig:casechocolatesinverse}
\end{figure}
\end{CodeChunk}

@@

(ref:casefifa-cap) Exploring the wages relative to skill measurements in
the FIFA 2020 data. Star offensive player (L. Messi) is the PI, and he
is compared with a top defensive player (V. van Dijk). The attribution
projection is shown on the left, and it can be seen that this
combination of variables produces a view where Messi has very high
predicted (and observed) wages. Defense (\texttt{def}) is the chosen
variable to vary. It starts very low, and Messi's predicted wages
decrease dramatically as its contribution increases (right plot). The
increased contribution in defense comes at the expense of offensive and
reaction skills. The interpretation is that Messi's high wages are most
attributable to his offensive and reaction skills, as initially provided
by the local explanation. (A video of the animated radial tour can be
found at \url{https://vimeo.com/666431163}.)

@@

\caption[(ref:casefifa-cap)]{(ref:casefifa-cap)}\label{fig:casefifa}
\end{figure}
\end{CodeChunk}

@@

(ref:caseames-cap) Exploring an observation with an extreme residual as
the PI in relation to an observation with an accurate prediction for a
similarly priced house in a random forest fit to the Ames housing data.
The local explanation indicates a sizable attribution to Lot Area
(\texttt{LtA}), while the CI has minimal attribution to this variable.
The PI has a higher predicted value than the CI in the attribution
projection. Reducing the contribution of Lot Area brings these two
prices in line. This suggests that if the model did not value Lot Area
so highly for this observation, then the observed sales price would be
quite similar. That is, the large residual is due to a lack of factoring
in the Lot Area for the prediction of PI's sales price. (A video showing
the animation is at \url{https://vimeo.com/666431134}.)

@@

\caption[(ref:caseames-cap)]{(ref:caseames-cap)}\label{fig:caseames}
\end{figure}
\end{CodeChunk}

>>>>

@@

\caption[Examining the SHAP values for a random forest model classifying Palmer penguin species]{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a) and the dashed vertical line in (b). The radial view shows varying the contribution of `fl` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at https://vimeo.com/666431172.)}\label{fig:casepenguins}
\end{figure}
\end{CodeChunk}

@@

@@

\caption[Checking what is learned from the cheem viewer]{Checking what is learned from the cheem viewer. This is a plot of flipper length (`fl`) and bill length (`bl`), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small `fl` length which is why it is confused with a Chinstrap.}\label{fig:casepenguinsblfl}
\end{figure}
\end{CodeChunk}

@@

@@

\caption[Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green)]{Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this observation. So, the contribution of Sugar is varied---reducing it corresponds primarily with an increased magnitude from Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at https://vimeo.com/666431143.)}\label{fig:casechocolates}
\end{figure}
\end{CodeChunk}

@@

@@

\caption[Examining the local explanation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange)]{Examining the local explanation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed variable importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at https://vimeo.com/666431148.)}\label{fig:casechocolatesinverse}
\end{figure}
\end{CodeChunk}

@@

@@

\caption[Exploring the wages relative to skill measurements in the FIFA 2020 data]{Exploring the wages relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI, and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown on the left, and it can be seen that this combination of variables produces a view where Messi has very high predicted (and observed) wages. Defense (`def`) is the chosen variable to vary. It starts very low, and Messi's predicted wages decrease dramatically as its contribution increases (right plot). The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at https://vimeo.com/666431163.)}\label{fig:casefifa}
\end{figure}
\end{CodeChunk}

@@

@@

\caption[Exploring an observation with an extreme residual as the PI in relation to an observation with an accurate prediction for a similarly priced house in a random forest fit to the Ames housing data]{Exploring an observation with an extreme residual as the PI in relation to an observation with an accurate prediction for a similarly priced house in a random forest fit to the Ames housing data. The local explanation indicates a sizable attribution to Lot Area (`LtA`), while the CI has minimal attribution to this variable. The PI has a higher predicted value than the CI in the attribution projection. Reducing the contribution of Lot Area brings these two prices in line. This suggests that if the model did not value Lot Area so highly for this observation, then the observed sales price would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at https://vimeo.com/666431134.)}\label{fig:caseames}
\end{figure}
\end{CodeChunk}
